Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-06-10 paper/main.tex
fixing everything cycle related

\begin{definition}
Let $G=(V,E_G)$ be a DAG let and $H=(V,E_H)$ be a directed graph (possibly cyclic) 
such that $E_{G^+} \subseteq E_{H^+}$. Then $\mathcal{A}_H$
is called a \emph{$G$-compatible ancestral oracle}. Further, 
if $H=(V,E_H)$ is an acyclic supergraph of $G$, then 
$\mathcal{A}_H$ is called a \emph{acyclic $G$-compatible ancestral oracle} with respect to $G$.
\end{definition}

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{FixCycle}{FixCycle}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc FixCycles}($V, E, \mathcal{D}_G $)}{
		$R \gets \emptyset$ \; 
		\ForEach {$ x_i \rightarrow x_j $ on a cycle in $ (V,E) $}{
		\If {there exists a set $\mathbf{Z}\subseteq\mathbf{X}$ where $\mathcal{D}_G(x_i, x_j)=1$
}{
			$ R \gets R \cup \{ x_i \rightarrow x_j \} $ \;

@@

$\textsc{Expand}$ takes an initial list of edges and searches for 
any unconnected vertex pairs that are not connected but where the
d-separation oracle indicates a residual association not explained 
by other paths.
The parameter $B$ specifies a ``black list''
of vertex pairs that must not be connected, 
whereas $k$
can be used to limit the maximum amount of edges 
to be added by this procedure; the usefulness of the $B$ and $k$
parameters will become clear soon.

@@

@@

By combining Propositions~\ref{prop:strongexpand}, \ref{prop:weakexpand} and 
\ref{prop:prune}, we immediately obtain the following:

\begin{theorem}
Given a d-separation oracle $\mathcal{D}_G$ and a $G$-consistent 
ancestral oracle $\mathcal{A}_H$, 
$\textsc{Prune}(\textsc{Expand}((V,\emptyset),\mathcal{D}_G,\mathcal{A}_H,\emptyset,\infty),\mathcal{D}_G)=G.$
\end{theorem}

@@

\Fn{{\sc ExpertInLoop}($V,\mathcal{D}_G,\mathcal{A}_G$)}{
$E_p \gets \emptyset$ \tcc{Current edges} 
$B \gets \emptyset$ \tcc{Edges that were pruned} 
\Repeat{ $E=E_p$ }{
	$E \gets E_p$ \;
	$(V,E) \gets \textsc{Expand}(V,E,\mathcal{D}_G,\mathcal{A}_G,B,1)$ \;
	$(V,E_p) \gets \textsc{Prune}(V,E,\mathcal{D}_G)$ \;
	$B \gets B \cup \{ E_p \setminus E \} $
}
\Return{(V,E)}
}

@@

@@

>>>>

\begin{definition}
Let $G=(V,E_G)$ be a DAG let and $H=(V,E_H)$ be a directed graph (possibly cyclic) 
such that $E_{G^+} \subseteq E_{H^+}$. Then $\mathcal{A}_H$
is called a \emph{$G$-compatible ancestral oracle}. If $H$ is furthermore acyclic, then 
$\mathcal{A}_H$ is called a \emph{acyclic $G$-compatible ancestral oracle} with respect to $G$.
\end{definition}

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{FixCycle}{FixCycle}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc FixCycles}($V, E, \mathcal{D}_G $)}{
		$R \gets \emptyset$ \; 
		\ForEach {$ X \rightarrow Y $ on a cycle in $ (V,E) $}{
		\If {there exists a set $\mathbf{Z}\subseteq\mathbf{X}$ where $\mathcal{D}_G(X, Y, \mathbf{Z})=1$
}{
			$ R \gets R \cup \{ X \rightarrow Y \} $ \;

@@

$\textsc{Expand}$ takes an initial list of edges and searches for 
any unconnected vertex pairs that are not connected but where the
d-separation oracle indicates a residual association not explained 
by other paths.
The parameter $B$ specifies a ``black list''
of vertex pairs that must not be connected, which is important
to prevent edges that were removed from cycles to be added again 
and will have another important role in the overall algorithm.
In addition, $k$
can be used to limit the maximum amount of edges 
to be added by this procedure; it will become clear soon why
this is useful.

@@

When using experts that generate cyclic graphs, edges are pruned to 
break these cycles, which can lead to a more sparse result.

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a $G$-compatible ancestral oracle $\mathcal{A}_H$ where
$H$ contains at least one cycle,
let $\tilde{G}=\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_H, \emptyset, \infty)$
Then $H^+ \supsetneq \tilde{G} \supseteq G$. 
\label{prop:weakexpand:cyclic}
\end{proposition}

\begin{proof}
All edges in the skeleton of $G$ are added during $\textsc{Expand}$ with the 
correct orientation and cannot be removed by $\textsc{FixCycles}$. Therefore, the result
is a supergraph of $G$. If cycles are created during $\textsc{Expand}$, then every 
cycle contains at least one edge that is not the skeleton of $G$, otherwise $G$ itself
would be cyclic. At least one edge of every cycle is therefore removed by
 $\textsc{FixCycles}$, making the result acyclic again.
\end{proof}

@@

By combining Propositions~\ref{prop:strongexpand}, \ref{prop:weakexpand}, 
\ref{prop:weakexpand:cyclic} and 
\ref{prop:prune}, we immediately obtain the following:

\begin{theorem}
Given a d-separation oracle $\mathcal{D}_G$ and a $G$-consistent 
ancestral oracle $\mathcal{A}_H$, 
let $(V',E',B)=\textsc{Expand}((V,\emptyset),\mathcal{D}_G,\mathcal{A}_H,\emptyset,\infty)$.
Then $\textsc{Prune}(V',E',\mathcal{D}_G)=G.$
\end{theorem}

@@

\Fn{{\sc ExpertInLoop}($V,\mathcal{D}_G,\mathcal{A}_G$)}{
$E_p \gets \emptyset$ \tcc{Current edges} 
$B \gets \emptyset$ \tcc{Edges that were pruned or removed from cycle} 
\Repeat{ $E=E_p$ }{
	$E \gets E_p$ \;
	$(V,E,B) \gets \textsc{Expand}(V,E,\mathcal{D}_G,\mathcal{A}_G,B,1)$ \;
	$(V,E_p) \gets \textsc{Prune}(V,E,\mathcal{D}_G)$ \;
	$B \gets B \cup \{ E_p \setminus E \} $
}
\Return{(V,E)}
}

@@

We note that the general property of Algorithm~\ref{algo:expert} that 
each possible edge can be added and removed at most once, which guarantees
that the algorithm eventually terminates, remains valid even if the oracles make
arbitrary mistakes. This is crucial when using the algorithm in practice.

@@

In addition to prioritizing edges to add or remove, the ranking could also
be used to break cycles when CI tests fail to identify the edges that are not
in the skeleton during the procedure  $\textsc{FixCycle}$, which could happen
due to finite sample effects. In that case, removing edges that correspond to 
weak associations may be preferable to keeping the cycles in the model.
